A Systematic Literature Review on Extensions to Role-Based Access Control

### 2013-05-01 paper/sections/abstract.tex
address comments
John's comments

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's by the National Institute of Standards and Technology (NIST),
computing in new domains has led to proposals for extensions to the RBAC standard model; for example adding context around permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC standard model, and in understanding
how extensions to RBAC are evaluated by providing an assessment of the state of extensions to the RBAC standard model. We accomplished this through: 
establishment of a set of extension categories, an examination of the state of the art in evaluations of extension models and a breakdown of the 
motivations that have led to extension of the RBAC standard model.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that yielded 1716 papers, of which 28 were deemed primary sources for inclusion as extensions to the RBAC standard model. 
\\
\\
\noindent
{\bf Results}: Our results show that extensions to RBAC can be classified under eight categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial and Temporal. We identified only
eight of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired extensions were the medical domain with nine of the 28 models, enterprise workflows with five of 28, and mobile computing with five of 28.
\\
\\
\noindent
{\bf Conclusions}: When examined as an entire group, the eight categories can be aligned under the single category of context. 
Our literature review showed that the state of evaluation of RBAC model evaluation is limited with one
paper doing an evaluation comparing itself to the core RBAC model, eight presenting example 
scenarios of their model in action, and 12 of the 28 models lacking any identifiable evaluation. The landscape of extensions
to the RBAC standard model serves as a basis from which a new, and vetted, extended standard RBAC model should be derived.

>>>>

\noindent
{\bf Context}: Since the National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) standard using RBAC reference models (RBAC$_{m}$ in short) in the late 1990's, computing in new domains has led to proposals for extensions to the RBAC$_{m}$; for example adding context around permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC$_{m}$, and in understanding
how extensions to RBAC$_{m}$ are evaluated by providing an assessment of the state of extensions to the RBAC$_{m}$. We accomplished this through: 
establishment of a set of extension categories, an examination of the state of the art in evaluations of extension models, and a breakdown of the 
motivations that have led to extension of the RBAC$_{m}$.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that yielded 1,716 papers, of which 28 were deemed primary sources for inclusion as extensions to the RBAC$_{m}$. 
\\
\\
\noindent
{\bf Results}: Our results show that extensions to the RBAC$_{m}$ can be classified under eight categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. We identified only
eight of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired extensions were the medical domain with nine of the 28 models, enterprise workflows with five of 28, and mobile computing with five of 28.
\\
\\
\noindent
{\bf Conclusions}: When examined as an entire group, the eight categories can be aligned under the single category of context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with one
paper doing an evaluation comparing itself to the RBAC$_{m}$, eight presenting example 
scenarios of their model in action, and 12 of the 28 models lacking an evaluation that we could discern. The landscape of extensions
to the RBAC$_{m}$ serves as a basis from which a new, and vetted, extended RBAC reference models should be derived.

### 2013-05-01 paper/sections/analysis.tex
address comments
John's comments

RQ2: What are the motivations behind extensions to RBAC?

@@

We first describe motivations described in 28 papers surveyed.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The RBAC standard needs additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC standard does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
\item The RBAC standard does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item The RBAC standard does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item The RBAC standard has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\item The RBAC standard needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC standard needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
is introduced. 
\item RBAC model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

@@

\begin{itemize}
	\item Addition: The extension model incorporates additional elements with regards to context/constraints and their relations with existing elements of the RBAC standard. Models in this category do not change the RBAC standard model. We found 23 papers in this category.
	\item Modification: The extension model incorporates modification of existing features in the RBAC standard. For example, changes of attribute relations such as role hierarchy are to show modification. For models in this category, changes occur within the the RBAC standard model. We found one paper in this category.
	\item Combination: The extension model is combination of RBAC and another access control model, which uses model-specific attributes and their relations such as task, team, purpose, organizational roles. For this combination, new association between attributes of two models should be introduced. Models in this category require a new association between attributes of different access control models to work together. We found four papers in this category.
\end{itemize}

@@

The RBAC standard provides a formal access model, which is an abstraction on top of authorization based on user role assignment (UA) and role
permission assignment (PA). Since this simple model may not provide a fine-grained access control for sophisticated
security mechanism, a variety of extended RBAC models have been proposed over the years to meet
security requirements. When administrators design a model, it is important to capture an important abstraction to help the model to be enforced in a system.

The RBAC standard presents four entities: roles, sessions, users, and permissions, and their relations. While this model is considered fundamental in any RBAC systems, this model can be extended to be applicable for emerging applications such as healthcare and mobile devices. 
In such environments, the RBAC standard needs additional constraints and dynamic context to support dynamic environments such as spatial and location. For this category, more than 80\% of papers are classified into ``Addition'' category. Besides, one paper~\cite{alam06:constraint} describes modification of existing relationship among elements to support partial inheritance. The RBAC standard is shown to be combined with other access control models such as team-based access control and task-based access control~\cite{zhou2007team, oh2003task}.

@@

The RBAC standard was designed with enterprises in mind such that when practitioners implemented RBAC into their systems
there would be a reasonable assurance being based off a well thought out model.  As extensions to the standard model
come along, thought and time should be given to how features and nuances of their models may impact implementation
in order to achieve the same goals as the original standard.  The primary sources should a significant lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as four papers did, should be
a prototype implementation of the model for review by both practitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

@@

Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. Eight models presented example scenarios
and how application of their model would apply and resolve the situation.  Six of the models provided some form of performance
or complexity analysis of their model.  This included graphs of the model's time to determine authorization as the number of entities
grew, and the size of the role space for the extension model compared to standard RBAC. Four models provided mathematical descriptions
and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation method was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
or show how the standard model is deficient in tackling said scenarios.

@@

When proposing an access control model, providing an evaluation of the model is a key component in establishing the validity of the model. 
Further, in the case of extensions to the RBAC standard model, the model should be accompanied by validation of the model as a stand-alone
access control model and in comparison to the model upon which the enhancements are being made. The results show that robust evaluations of
extension models are lacking.

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. In rare cases, five primary sources, the model 
authors provide some discussion of how the RBAC standard model is deficient in tackling the scenario. In rare cases, 
one paper from our results, a case study is performed to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC standard
model, developers may be left implementing a more complex model to address their requirements when the standard model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

When looking for an enterprise ready access control mechanism, developers must balance usability with security. 
Two of the primary sources examined provided time-based performance analysis of their extension model compared
to the RBAC standard model. This inclusion of time analysis provides some assurances to developers that any non-functional
requirements surrounding time to compute authorizations compete or beat the standard model. Further, four of the models provided
some form of complexity of their model. This complexity analysis plays a key role in the management of the access control mechanism
over the course of the models implementation lifetime. As the number of roles, users and additional entities grows, developers will 
need to ensure non-functional requirements are met that deal with the ability for a system administrator to effectively manage these
entities.

@@

RQ5: What domains have extensions to RBAC been created for?

@@

The RBAC standard has been re-enforced by the economic impact that standardization has had on enterprises needing to apply access control.  The
inclusion of extension models targeted at the enterprise workflow domain is indicative of the expansion of requirements for enterprises. Developers
and researchers should take care when looking at extension models designed to addresses the newer requirements of enterprise workflows in order to
achieve the same economic implementation and maintainability benefits the original standard model presents.

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true (Permit) or false (Deny) based on predicates, propositional logic is sufficient to describe key ideas and definitions. We found that 28 papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps reduce the administrative complexity of associations such as user- role associations. An alternative logic to describe RBAC is first-order logic. This logic is sufficient to describe RBAC and extended RBAC. Moreover, this logic supports for concise and elegant formulation of the RBAC model and its relation.  First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.

>>>>

RQ2: What are the motivations behind RBAC extensions?

@@

We first describe motivations described in the 28 papers surveyed.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The RBAC$_{m}$ needs additional contextual information and constraints to develop fine-grained policies.
\item The RBAC$_{m}$ does not incorporate context. Therefore, RBAC belongs to the static access control model, which may not capture changes in environments.
\item The RBAC$_{m}$ does not support various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item The RBAC$_{m}$ does not provide an abstraction for additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item The RBAC$_{m}$ has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
%\item The RBAC standard needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item The RBAC$_{m}$ needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new associations between attributes
are introduced. 
\item The RBAC$_{m}$ needs to improve existing features such as delegation and role hierarchy to provide fine-grained access control. 
\end{itemize}

@@

\begin{itemize}
	\item Addition: The extension model incorporates additional elements with regards to context/constraints and their relations with existing elements of the RBAC standard. Models in this category do not change the RBAC$_{m}$. We found 23 papers in this category.
	\item Modification: The extension model incorporates modification of existing features in the RBAC$_{m}$. For example, changes of attribute relations such as role hierarchy fall into this modification category. For models in this category, changes occur within the RBAC$_{m}$. We found one paper in this category.
	\item Combination: The extension model is a combination of RBAC$_{m}$ and another access control model, which uses model-specific attributes and their relations such as task, team, purpose, and organizational roles. For this combination, new associations between attributes of two models should be introduced. Models in this category require new associations between attributes of different access control models to work together. We found four papers in this category.
\end{itemize}

@@

The RBAC standard provides RBAC$_{m}$, which is an abstraction on top of authorization based on $UA$ and $PA$. Since RBAC$_{m}$ may not provide a fine-grained access control for a sophisticated
security mechanism, a variety of extended RBAC models have been proposed over the years to meet
security requirements. When administrators design a model, it is important to capture an important abstraction to help the model to be enforced in a system.

RBAC$_{m}$ presents four entities: roles, sessions, users, and permissions, and their relations. While RBAC$_{m}$ is considered fundamental in any RBAC systems, RBAC$_{m}$ can be extended to be applicable for emerging applications such as healthcare and mobile devices. 
In such environments, the RBAC standard needs additional constraints and dynamic context to support dynamic environments such as spatial and location. For this category, more than 80\% of papers are classified into ``Addition'' category. Besides, one paper~\cite{alam06:constraint} describes modification of existing relationship among elements to support partial inheritance. The RBAC standard is shown to be combined with other access control models such as team-based access control and task-based access control~\cite{zhou2007team, oh2003task}.

@@

The RBAC standard was designed with enterprises in mind such that when practitioners implemented RBAC into their systems
there would be a reasonable assurance being based off a well thought out model.  As extensions to the RBAC$_{m}$
come along, thought and time should be given to how features and nuances of their models may impact implementation
in order to achieve the same goals as the original standard.  The primary sources should a significant lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as four papers did, should be
a prototype implementation of the model for review by both practitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

@@

Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. Eight models presented example scenarios
and how application of their model would apply and resolve the situation.  Six of the models provided some form of performance
or complexity analysis of their model.  This included graphs of the model's time to determine authorization as the number of entities
grew, and the size of the role space for the extension model compared to standard RBAC. Four models provided mathematical descriptions
and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation method was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
or show how the RBAC$_{m}$ is deficient in tackling said scenarios.

@@

When proposing an access control model, providing an evaluation of the model is a key component in establishing the validity of the model. 
Further, in the case of extensions to the RBAC$_{m}$, the model should be accompanied by validation of the model as a stand-alone
access control model and in comparison to the model upon which the enhancements are being made. The results show that robust evaluations of
extension models are lacking.

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. In rare cases, five primary sources, the model 
authors provide some discussion of how the RBAC$_{m}$ is deficient in tackling the scenario. In rare cases, 
one paper from our results, a case study is performed to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC$_{m}$, developers may be left implementing a more complex model to address their requirements when the RBAC$_{m}$ would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

When looking for an enterprise ready access control mechanism, developers must balance usability with security. 
Two of the primary sources examined provided time-based performance analysis of their extension model compared
to the RBAC$_{m}$. This inclusion of time analysis provides some assurances to developers that any non-functional
requirements surrounding time to compute authorizations compete or beat the RBAC$_{m}$. Further, four of the models provided
some form of complexity of their model. This complexity analysis plays a key role in the management of the access control mechanism
over the course of the models implementation lifetime. As the number of roles, users and additional entities grows, developers will 
need to ensure non-functional requirements are met that deal with the ability for a system administrator to effectively manage these
entities.

@@

RQ5: What domains have RBAC extensions been created for?

@@

The RBAC standard has been re-enforced by the economic impact that standardization has had on enterprises needing to apply access control.  The
inclusion of extension models targeted at the enterprise workflow domain is indicative of the expansion of requirements for enterprises. Developers
and researchers should take care when looking at extension models designed to addresses the newer requirements of enterprise workflows in order to
achieve the same economic implementation and maintainability benefits the RBAC$_{m}$ presents.

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true (Permit) or false (Deny) based on predicates, propositional logic is sufficient to describe key ideas and definitions. We found that 28 papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps reduce the administrative complexity of associations such as user- role associations. An alternative logic to describe RBAC is first-order logic. This logic is sufficient to describe RBAC and extended RBAC. Moreover, this logic supports for concise and elegant formulation of the RBAC$_{m}$ and its relation.  First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.

### 2013-05-01 paper/sections/categorization.tex
address comments
John's comments

During the paper reading phase, the authors noted that categorical themes were present among the extensions. 
For example, the paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} brings in the notion of privacy directly within the title of the paper and the name of their model. 
Some papers presented this direct pronouncement of a theme, while others were less obvious. 
Thus, we developed a set of guidelines to aide in determining a set of extension categories based upon observations during the paper reading phase. 
In developing these guidelines, we attempted to define each category by a single noun-phrase descriptor.
The guidelines were:

@@

If we take the previous example above, the paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} contained ``privacy'' in
the title and in the name of the model leading to the creation of the Privacy category and the subsequent placement of the paper under that category.
By comparison, the paper ``An extended RBAC model based on granular logic'' \cite{jian2008extended} does not contain a direct categorization in the title or model name, but in reading the body of the paper the authors assessed that this paper was context based.

We offer definitions for each of the eight observed categories based on the data extracted from the primary sources and the English definitions for each noun-phrase. 
Some category descriptors contain abbreviations in parenthesis that match the shortened name found within Table \ref{tab:categorization}.

@@

  \item \textbf{Context}: The extension model integrates contextual information into the RBAC standard model. Context is defined as a user's current state and environment information (e.g., location, time, system resources, network state, network security configuration, etc) which may affect user's access privileges.

  \item \textbf{Constraint (Const)}: The extension model provides conditional restrictions on permissions of given roles. The constraint is either static or dynamic. For example, a doctor can modify any medical record for which the doctor is assigned as the designated primary care physician. This example describes a doctor's permission with conditional restriction - ``only when the doctor is assigned as the designated primary care physician can the doctor modify a particular medical record''.

@@

  \item \textbf{Privacy (Priv)}: The extension model provides entities and mechanisms to describe privacy policies, which are legal statements or documents about disclose and management of personally identifiable information such as name, address, data of birth, etc.

  \item \textbf{Task}: The extension model provides task elements which are associated with permissions and roles. A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are first associated with tasks, which are associated permissions. For example, the employee role is associated with a task, which is to write a report. Then, this task is associated with a permission element.

  \item \textbf{Spatio-Temporal}: The extension model combines spatial (location-based) and temporal (time-based) constraints in specifying access control policies. For example, specific locations are enforced while a role is permitted to conduct an action from 8 am to 5 pm.

  \item \textbf{Spatial}: The extension model provides spatial constraints, location-based constraints in specifying access
	control policies. For example, in organizations, locations are enforced while a
	specific role is permitted to conduct an action. Consider that a part-time employee works only in a specific location.
	In such cases, the part-time employee role should access required resources only when the user is in the location. 
	Spatial constraints can integrate with roles, user-role assignments, or role-permission assignments.

  \item \textbf{Temporal (Temp)}:  The extension model provides temporal constraints, time-based constraints in specifying access
	control policies. For example, in organizations, periodic temporal durations are enforced while a
	specific role is permitted to conduct an action. Consider that a part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should access required resources during the interval. 
	Temporal constraints can integrate with roles, user-role assignments, or role-permission assignments.

@@

The 28 primary sources produced a set of eight hierarchical categories. Table \ref{tab:categorization} summarizes each primary source under their designated
category and further, displays the perceived hierarchy of the categories. 
The Spatial and Temporal categories were treated as subsets of the broader category of Spatio-Temporal since this category encompasses them individually and
the Spatio-Temporal category contained more primary sources that the Spatial or Temporal categories alone.

When looking across all categories, the authors noted that each category added some new features on top of the standard RBAC model that were domain specific.
These domain specific features were under the surface adding contextual relationships between the core user, permission and role entities.  Thus, the authors
concluded that all categories stemmed from the context category, of which some primary sources were already deemed direct members.

>>>>

During the paper reading phase, we identify categorical themes, which are
central topic or context of RBAC extensions. 
For example, the paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} brings in the notion of privacy explicitly within the title of the paper and the name of their model. 
Some papers presented this direct pronouncement of theme of RBAC extensions, while others were less obvious. 
Thus, we developed a set of guidelines to aide in determining a set of eight categories based upon observations during the paper reading phase. 
In developing these guidelines, we defined each category by a single noun-phrase descriptor.
The guidelines were:

@@

The previous example paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} contained ``privacy'' in
the title and in the name of the model leading to the creation of the Privacy category and the subsequent placement of the paper under that category.
By comparison, the paper ``An extended RBAC model based on granular logic'' \cite{jian2008extended} does not contain a direct categorization in the title or model name. However, in reading the body of the paper, we determined that this paper discussed RBAC extension based on context. 
We offer definitions for each of the eight observed categories based on the data extracted from the primary sources and the English definitions for each noun-phrase. 
Some category descriptors contain abbreviations in parenthesis that match the shortened name found within Table \ref{tab:categorization}.

@@

  \item \textbf{Context}: The extension model integrates contextual information into the RBAC standard model. Context is defined as a user's current state and environment (e.g., location, time, system resources, network state, network security configuration, etc) which may affect the user's access privileges.

  \item \textbf{Constraint (Const)}: The extension model provides conditional restrictions on permissions of given roles. The constraint is either static or dynamic. For example, a doctor may modify any medical record for which the doctor is assigned as the designated primary care physician. This example describes a doctor's permission with a conditional restriction - ``only when the doctor is assigned as the designated primary care physician may modify a particular medical record''.

@@

  \item \textbf{Privacy (Priv)}: The extension model provides entities and mechanisms to describe privacy policies, which are legal statements or documents about disclosure or management of personally identifiable information such as name, address, data of birth, etc.

  \item \textbf{Task}: The extension model provides task elements which are associated with permissions and roles. A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are associated with tasks that are associated with permissions. For example, the employee role is associated with a task, for example, to write a report. Then, this task is associated with a permission.

  \item \textbf{Spatio-Temporal}: The extension model combines spatial (location-based) and temporal (time-based) constraints in specifying access control policies. For example, specific locations permit roles to conduct actions from 8:00 a.m. to 5:00 p.m.

  \item \textbf{Spatial}: The extension model provides spatial constraints, location-based constraints in specifying access
	control policies. For example, in organizations, locations are enforced while a
	specific role is permitted to conduct an action. Consider that a part-time employee works only in a specific location.
	In such cases, the part-time employee role should access required resources only when the user is in that location. 
	Spatial constraints can integrate with roles, user-role assignments, or role-permission assignments.

  \item \textbf{Temporal (Temp)}:  The extension model provides temporal constraints, time-based constraints in specifying access
	control policies. For example, in organizations, periodic temporal durations are enforced while a
	specific role is permitted to conduct an action. Consider that a part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should be allowed to access required resources during the interval. 
	Temporal constraints can integrate with roles, user-role assignments, or role-permission assignments.

@@

The 28 primary sources produced a set of eight hierarchical categories. Table \ref{tab:categorization} summarizes each primary source under their designated
category and furthermore, displays the perceived hierarchy of the categories. 
The Spatial and Temporal categories were treated as subsets of the broader category of Spatio-Temporal since this category encompasses them individually and
the Spatio-Temporal category contained more primary sources that the Spatial or Temporal categories alone.

When looking across all categories, the authors noted that each category added some new features on top of the standard RBAC model that were domain specific.
These domain specific features were under the surface these features adding contextual relationships between the core user, permission and role entities.  Thus, the authors
concluded that all categories stemmed from the context category, of which some primary sources were already deemed direct members.

### 2013-05-01 paper/sections/conclusion.tex
fixing after spell check and grammar check based on word
spell check and grammar check based on word
address comments
John's comments

The extensions to RBAC were revealed to fall into a number of categorizations with Organization, Privacy, Resource, Task, Spatio-Temporal, Spatial and Temporal falling under the general category of context.
The categories each had properties specific to their implementation, but were seen to generalize to being specialized instances of context tailored to the entities or actions the categories covered.
A number of domains were idenitfied as being the motivations behind needing extensions to the core RBAC model.  The domains, such as healthcare, presenting new challenges the previous models were not
required to design for.  Our literature review showed that the state of RBAC model evaluation needs focused from the research community given most model evaluations seen within the papers were
based on hypothetical situations with little to no case studies or implementations in practice.

>>>>

The RBAC extensions were revealed to fall into a number of categorizations with Organization, Privacy, Resource, Task, Spatio-Temporal, Spatial, and Temporal falling under the general category of context.
The categories each had properties specific to their implementation, but were seen to generalize to being specialized instances of context tailored to the entities or actions the categories covered.
A number of domains were identified as being the motivations behind needing extensions to the core RBAC model.  The domains, such as healthcare, presenting new challenges the previous models were not
required to design for.  Our literature review showed that the state of RBAC model evaluation needs focused from the research community given most model evaluations seen within the papers were
based on hypothetical situations with little to no case studies or implementations in practice.

### 2013-05-01 paper/sections/core_rbac.tex
address comments
John's comments

\section{Role Based Access Control Standard} \label{sec:core-rbac}

RBAC provides effective and efficient permissions management for operations, especially when sharing resources within an organization.
Prior to the creation of the NIST RBAC standard, no general agreement on the definition 
of RBAC existed among practitioners or within the research community. 
Without a unified definition of RBAC, software developers described similar concepts and features of RBAC models using different terminology. 
This lack of consistent terminology was shown to slow the implementation of RBAC \cite{o20102010}.  
Moreover, in cases where organizations were concerned with adopting RBAC,
evaluation and comparison of RBAC technologies developed by different vendors was difficult.
NIST, in collaboration with industry and academics, worked on defining a set of consensus RBAC concepts and terminology and proposed a standard for 
RBAC that addressed these cost and interoperability issues by developing a common definition that can be used across different vendors.

NIST's work can directly benefit organizations by lowering cost of early phase R\&D and the implementation of RBAC.
Since the RBAC standard was first introduced, a 2010 report by RTI International showed that the the rate of RBAC adoption has rapidly grown over recent years \cite{o20102010}. 
The analysts estimate that, by the end of 2010, at least a part of permission of systems use RBAC for more than 50\% of IT users at organizations, which
hire more than 500 employees. For economic estimate analysis, the analysts estimate that RBAC technology has generated \$6.1 billion in net economic benefits to industry, of which \$1.1 billion is attributable to NIST RBAC standard work.

The NIST RBAC standard model proposed by Ferraiolo et al. \cite{ferraiolo} and later adopted as the official standard for RBAC by the INCITS includes three levels of RBAC: core RBAC, hierarchical RBAC, and constrained RBAC.
Since the basis for our review is extensions to the RBAC standard model, we describe the core model, associated entities and other terminology encountered across the space of our review. Then, we describe hierarchical RBAC and constrained RBAC, which are developed by incorporating new features into the core RBAC.

@@

The four entities of the standard RBAC model are:

@@

In RBAC, a user can exercise a permission only if the user is assigned to a role.
In addition to the four basic components, two functions are defined:
the user role assignment (UA) and the role and permission assignment (PA) functions.
User role assignment models assignment of users to roles.
Permission role assignment models assignment of permissions to roles.

\begin{figure}[ht]
    \centering
        \includegraphics[width=4.0in]{sections/core-model.png}
    \caption{\label{fig:overview}RBAC Core Diagram in NIST RBAC model \cite{ferraiolokuhn}.}
\end{figure}

Figure~\ref{fig:overview} presents an overview of core RBAC model diagram where elements and their relations are described.
Let "USERS", "ROLES", "OBS", "OPS", "PRM", and "SESSIONS" denote users, roles, objects, operations, permissions, and sessions, respectively.
Permissions are associated with possible users' pre-defined operation on an object (e.g., execute a file).
Note that, at user or role activation, a session associated with user or role is established.

@@

Hierarchical RBAC model adds role hierarchies (RH) feature to core RBAC model.
This model incorporates a structure of roles in an organization using inheritance relationship among attributes such as roles.
The role structure in an organization may use
a role $r_1$, who inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of a employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments to a user. Formally, role inheritance relation is shown as \textit{RH} $\subseteq$ \textit{ROLES} $\times$ \textit{ROLES} describing many-to-many mapping role inheritance relation. 
General role hierarchies can be extended to use the concept of multiple inheritance where
a role $r_1$, who inherits all permissions from more than one roles.

@@

Constrained RBAC incorporates static and dynamic separation of duty relations to the RBAC model. Separation of 
duty relations enforces conflict of interest among roles. This model defines two types of separation of duty relations; static and dynamic.

\begin{itemize}
	\item Static Separation of Duty (SSoD): SSoD restricts the conflicting-role assignments statically that are associated with a user. On situations
	where multiple roles can be associated with a single user and roles $Role_A$ and $Role_B$ are conflicting each other, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$. SSoD is known to be too rigid for practical use in cases where a user should have permissions when a user is assigned to both $Role_A$ and $Role_B$.
	\item Dynamic Separation of Duty (DSD): Dynamic SoD is known to be
more flexible than SSD. DSD restricts the conflicting-role assignments dynamically that are associated with a user. On situations
	where multiple roles can be associated with a single user and, roles $Role_A$ and $Role_B$ are conflicting each other dynamically, no permission is given to a user.	
\end{itemize}

>>>>

\section{Role-Based Access Control Standard} \label{sec:core-rbac}

RBAC provides effective and efficient permissions management for operations, especially when sharing resources within an organization.
Prior to the creation of the NIST RBAC standard, no general agreement on the definition 
of RBAC existed among practitioners or within the research community. 
Without a unified definition of RBAC, software developers described similar concepts and features of RBAC models using different terminology. 
This lack of consistent terminology was shown to slow the implementation of RBAC~\cite{o20102010}.  
Moreover, in cases where organizations were concerned with adopting RBAC,
evaluation and comparison of RBAC technologies developed by different vendors was difficult.
NIST, in collaboration with industry and academics, worked on defining a set of consensus RBAC concepts and terminology and proposed a standard for 
RBAC that addressed these cost and interoperability issues by developing a common definition that can be used across different vendors.

NIST's work can directly benefit organizations by lowering cost of early phase Research and Development, and implementation of RBAC.
Since the RBAC standard was first introduced, a 2010 report by RTI International showed that the the rate of RBAC adoption has rapidly grown over recent years \cite{o20102010}. 
The analysts estimate that by the end of 2010 at least a portion of permission of systems will use RBAC for more than half of users at organizations with more than 500 employees. The analysts estimate that RBAC technology has generated \$6.1 billion in net economic benefits to industry where NIST RBAC standard work saved \$1.1 billion.

The NIST RBAC standard proposed by Ferraiolo et al. \cite{ferraiolo} and later adopted as the official standard for RBAC by the INCITS includes three components of RBAC: core RBAC, hierarchical RBAC, and constrained RBAC. Each component includes a RBAC reference model. RBAC$_{m}$ includes RBAC reference models of these three components. We first describe the core RBAC, associated entities and other terminology encountered across the space of our review. We next describe hierarchical RBAC and constrained RBAC, which are developed by incorporating new features into the core RBAC.

@@

The four entities of the core RBAC reference model are:

@@

In RBAC, a user can exercise a permission only if the user is assigned to a role.
In addition to the four basic entities, two functions are defined:
user assignment ($UA$) and permission assignment ($PA$) functions.
$UA$ assignment of users to roles.
$PA$ assignment of permissions to roles.

\begin{figure}[ht]
    \centering
        \includegraphics[width=4.0in]{sections/core-model.png}
    \caption{\label{fig:overview}Diagram in Core RBAC reference model\cite{ferraiolokuhn}.}
\end{figure}

Figure~\ref{fig:overview} presents an overview of core RBAC reference model diagram where elements and their relations are described.
Let "USERS", "ROLES", "OBS", "OPS", "PRM", and "SESSIONS" denote users, roles, objects, operations, permissions, and sessions, respectively.
Permissions are associated with possible users' pre-defined operation on an object (e.g., execute a file).
Note that, at user or role activation, a session associated with user or role is established.

@@

The hierarchical RBAC reference model adds role hierarchies ($RH$) feature to the core RBAC reference model.
$RH$ incorporates a structure of roles in an organization using inheritance relationships among attributes such as roles.
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of a employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments for a user. Formally, the role inheritance relation is shown as \textit{RH} $\subseteq$ \textit{ROLES} $\times$ \textit{ROLES} describing the many-to-many mapping role inheritance relation. 
General role hierarchies can be extended to use the concept of multiple inheritances where
$r_1$ inherits all permissions from more than one roles.

@@

The constrained RBAC reference model incorporates separation of duty relations to  the core or the hierarchical RBAC reference. Separation of 
duty relations enforces conflict of interest among roles. This model defines two types of separation of duty relations; static and dynamic.

\begin{itemize}
	\item Static Separation of Duty (SSoD): SSoD restricts the conflicting-roles, which can be assigned to a single user statically. Consider that roles $Role_A$ and $Role_B$ are conflicting with each other. On situations
	where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ statically. SSoD is known to be too rigid for practical use in cases where a user should have permissions when a user is assigned to both $Role_A$ and $Role_B$.
	\item Dynamic Separation of Duty (DSD): Dynamic SoD is known to be
more flexible than SSoD. DSD restricts the conflicting-role assignments dynamically that are associated with a user. Consider that roles $Role_A$ and $Role_B$ are conflicting with each other. For the situations where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ dynamically.	
\end{itemize}
